Modelling state-switching behaviour in functional data with hidden Markov models

### 2020-09-04 Report/Sections/discussion.tex
Clean up report and delete old files
Update report- Evan Edits

Current FDA literature addresses dependence between curves either with multilevel models with random effects and no time component \citep{Chen:2012,Di:2009}, or with functional time series which model the curves as an evolving process \citep{Kokoszka:2018} and overlook the possibility that curves have several distinct ``types". However, many curve sequences exhibit both temporal dependence and a discrete number of types. One obvious example is the case study examined here, where one marine mammal dive serves a purpose such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}. Another example is accelerometer data in school children, where bouts of activity can be associated with a behaviour such as running or walking. \citep{Morris:2007}.

In this work, we deal with both temporal dependence between curves and distinct curve ``types" by using either an HMM or CarHMM. To help select the appropriate model between these two, lag plots can be used to test if there is significant auto-correlation between curves. A traditional HMM approach requires that practitioners take summary statistics of curves as emissions. However, HMM emissions can also be the entire curve, which is then modelled with a fine-scale parametric model. The parameters of this fine-scale model then depend upon the curve type. There are a wide range of possible fine-scale models, including another HMM, CarHMM, a spline-based approach similar to that of \citet{Langrock:2018}, and many more. Together, the two levels make up a hierarchical model which can account for processes taking place simultaneously at different time-scales.

%Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics when conditioned of the hidden state. However, many real-world processes require more complex models. Here, we presented a flexible framework that can account for complicated dependence structures within time-series data. Our framework is a collection of HMM models which can be combined together to form increasingly complex hierarchical models to match the complexity of particular problems faced by researchers.

%The CarHMM generalizes the HMM by explicitly modeling auto-correlation in the emission distributions of the HMM \citep{Lawler:2019}. CarHMMs also maintain the structure needed to evaluate the likelihood using the forward algorithm. In our Normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state.

When the fine-scale model is  an HMM, the resulting overall model is the HHMM as introduced by \citet{Barajas:2017}. \citet{Adam:2019} in particular use the HHMM to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California. This model can be inappropriate at very fine scales, as \citet{Adam:2019} find clear violations of conditional independence given hidden states and also observe periodic behaviour in the acceleration data. This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies.

%For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs. In particular, each hidden state of the coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.

%For processes such as these with very high sampling frequencies and/or intricate fine-scale structure,

%To deal with these issues, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood. For example, if the sampling rate of the fine-scale process is very high and clearly period, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.

To deal with these problematic fine-scale phenomena, we explore a moving-window transformation over the fine-scale observations. For periodic behaviour in particular, we use a DFT-based transformation. The effectiveness of this approach is evidenced by both our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale, and our simulation study, where the subdive decoding accuracy is by far the worst when omitting $\Ztwo$ (``wiggliness'') from the model. For simpler transformations, e.g. an average such as $\Zone$, either subject area experts or exploratory plots may suggest temporal correlation. In this case,  we suggest using a CarHMM instead of a traditional HMM. Our simulation study shows how ignoring this auto-correlation can result in underestimating standard errors and overestimating the variance of the observations.

We combine generalizations of HMMs with hierarchical structures and moving-window transformations to model animal behaviour as recorded from accelerometers. Time depth recorders and accelerometers are powerful tools to understand the diving behaviour and energetic requirements of NRKWs, and this is essential to understand their foraging behaviour and survival \citep{Williams:2009,Noren:2011}. However, biologging data require classifying the underlying activity \citep{Dot:2016} or dive type \citep{Hastie:2006} of the animal in order to effectively understand and estimate energy expenditure. The CarHHMM-DFT separates dives into different types on the coarse scale while describing the distribution of acceleration within each subdive state on the fine scale. This gives a deeper understanding of the animal's tri-axial movement, and thus its fine-scale energy expenditure \citep{Gleiss:2011,Qasem:2012}. Although our case study focuses on Northern resident killer whales, this model is likely applicable to many dive animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.

While we combine the HMMs of section \ref{sec:models} effectively for our case study, in general this practice should be done with care as there are drawbacks to using exclusively HMMs to build these hierarchical models. For example, HMMs require observations to be taken at regular intervals. Continuous-time methods such as the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019} and the continuous-time HMM (CTHMM) \citep{Liu:2015} are often based on stochastic differential equations and resolve this issue. However most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit. We show the appendix that the state-switching OU process of \citet{Michelot:2019} is equivalent to a CarHMM under certain conditions. This equivalence implies that a CarHMM is interpretable as a continuous-time model, but has the added advantage of relative computational efficiency.

When using HMMs to build complex models, one must consider how to effectively fit the data while avoiding over-fitting and at the same time ensuring computationally efficient parameter estimation. One way to temper model complexity is to assume that the fine-scale state parameters are the same for all coarse-scale states. Regardless, model complexity inevitably grows rapidly as more layers are added to the hierarchical structure. For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models. The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioural states. However, this model does not allow  ecologists  to study the joint relationship between the dive type and intra-dive behaviour.

%The simulation study also shows that the observed Fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases. One notable exception is the HHMM-DFT, which ignores auto-correlation and therefore underestimates the standard error of the parameters associated with acceleration. This can be dangerous in the field of ecology, as overconfident parameter estimates can result in overconfident conservation decisions. While the point estimates are generally good in this case study, this may not be the case in more general settings and for shorter time series in particular.

%We used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada. The CarHHMM-DFT was able to simultaneously distinguish three distinct subdive behaviors and two dive types. The DFT component proved useful in determining the subdive behaviour of the whale, as the mean of the emission distribution of $Z^{*(2)}$ for each subdive state was separated by an order of magnitude. Finally, the estimated auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and subdive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data. The introduction of the parameter $\phi^*$ also allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).

%Because traditional information criteria tend to overestimate the number of states in biological processes \citep{Pohle:2017}, the number of dive types and subdive behaviors was selected in an ad-hoc manner. There does appear to be some heterogeneity within dive types, and future work can be done to determine the optimal number of dive types and within-dive behaviors.

Our work provides a flexible framework to model functional time-series data using HMMs. Provided the model is not overly complex, a hierarchical model constructed using the proposed building blocks can be both flexible and easy to fit using maximum likelihood methods. We demonstrate the usefulness of such a model using an ecological example, where we use a complex HMM to classify the coarse and fine scale diving behaviour of a killer whale. In addition, as complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology to the analysis of a wide range of time-series data in a variety of fields.

>>>>

Current FDA literature addresses dependence between curves either with multilevel models with random effects and no time component at all \citep{Chen:2012,Di:2009}, or with functional time series which model the curves as an evolving process \citep{Kokoszka:2018} which overlook the possibility that curves have several distinct ``types". However, many curve sequences exhibit both temporal dependence and a discrete number of types. 
%One obvious example is the case study examined here, where one marine mammal dive serves a distinct purpose such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}. Another example is accelerometer data in school children, where bouts of activity can be associated with a behaviour such as running or walking. \citep{Morris:2007}. In this work, we deal with both temporal dependence between curves and distinct curve ``types" by using either an HMM or CarHMM. To help select the appropriate model between these two, lag plots can be used to test if there is significant auto-correlation between curves.

%When used traditionally, each of these models require that practitioners take summary statistics of curves as emissions which may lack important information about the structure of the curve. However, HMM emissions can also be viewed as the entire curve, where the emission distribution of that curve is described by yet another fine-scale model. There are a wide range of possible fine-scale models, including, but not limited to, another HMM, a CarHMM, a Poisson process, or a spline-based approach similar to that of \citet{Langrock:2018}. The parameters of this fine-scale model then depend upon the curve type as determined by the coarse-scale model. Together, these coarse- and fine-scale models make up a hierarchical model which can account for simultaneous processes taking place at different time-scales.

%One interesting class of possible fine-scale models is the set of continuous time methods such as the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019} and the continuous-time HMM (CTHMM) \citep{Liu:2015}, which are often based on stochastic differential equations. Unfortunately, most continuous-time models require MCMC algorithms to perform inference and can be slow to fit. However, we show the appendix that the state-switching OU process of \citet{Michelot:2019} is equivalent to a CarHMM under certain conditions. This equivalence implies that a CarHMM is interpretable as a continuous-time model, but has the added advantage of relative computational efficiency.

%Another fine-scale model that we investigate in detail here is yet another HMM. If the fine-scale model is an HMM, then the resulting overall model is just the HHMM as introduced by \citet{Barajas:2017}. \citet{Adam:2019} in particular use the HHMM to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California. This model can be inappropriate at very fine scales, as \citet{Adam:2019} find that subsequent fine-scale observations are not independent when conditioned on the hidden states due to periodic behaviour in the acceleration data. This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies. To deal with these problematic fine-scale phenomena, we explore a moving-window transformation over the fine-scale observations. Since our case study data clearly exhibits periodic behaviour, we specifically use a DFT-based transformation. The effectiveness of this approach is evidenced by both our case study, where the CarHHMM without a DFT transformation visually fails to recognize fluking behaviour in a killer whale, and our simulation study, where the average accuracy associated with the subdive state is by far the worst when omitting $\Ztwo_{t,t^*}$ (``wiggliness'') from the model. After transforming the fine-scale observations, either subject area experts or exploratory analysis may suggest a large degree of temporal correlation. In this case, we suggest using a CarHMM instead of a traditional HMM to account for the auto-correlation in the transformed fine-scale observations. Our simulation study shows how ignoring this auto-correlation can result in underestimating standard errors and overestimating the variance of the observations.

%When using HMMs to build complex hierarchical models, it is important to build a model that is complex enough to fit the data but not so complex that it over-fits the data or is computationally inefficient. Model complexity inevitably grows rapidly as more layers are added to the hierarchical structure. For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models. The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioural states. However, this model does not allow ecologists to study the joint relationship between the dive type and intra-dive behaviour. One way to temper model complexity is to reduce the number of parameters within a given model structure. In particular, we assume that the fine-scale state parameters are the same for all coarse-scale states for all models in both our case study and simulation study.

%To illustrate the potential applications of this method, we model the behaviour of a northern resident killer whale animal behaviour using data collected from time depth recorders and accelerometers. Understanding the diving behaviour and energetic requirements of NRKWs is essential to understand their foraging behaviour and survival \citep{Williams:2009,Noren:2011}. However, it is necessary to classify the underlying activity \citep{Dot:2016} or dive type \citep{Hastie:2006} of an animal in order to effectively understand and estimate its energy expenditure from biologging data. The CarHMMM-DFT model uses coarse-scale data to separate dives into different types while also describing the distribution of acceleration within each subdive state on the fine scale.

%Our work provides a flexible framework to model functional time-series data using HMMs.
We deal with temporal dependence between curves and distinct curve ``types" by using either an HMM or CarHMM.
Provided the model is not overly complex, a hierarchical model constructed using the proposed building blocks can be both flexible and easy to fit using maximum likelihood methods. We demonstrate the usefulness of such a model using an ecological example, where we use a complex HMM to classify the coarse- and fine-scale diving behaviour of a killer whale. This gives a deeper understanding of the animal's tri-axial movement and thus its fine-scale energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both NRKWs and their prey \citep{Noren:2011}. Although our case study focuses on northern resident killer whales, this model is likely applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}. In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time-series data in a variety of fields.

### 2020-09-04 Report/Sections/intro.tex
Clean up report and delete old files
Update report- Evan Edits

Biologging technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}. In particular, the study of animal movement has been transformed by tracking devices that collect biologging data \citep{Borger:2020}. Tags can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. These data contain a wealth of knowledge about human and animal behaviour, but parsing this vast amount of information is a challenge for statisticians.

%Many statistical techniques have been developed and applied to these types of longitudinal or functional data.  
%Some techniques, such as time series, are highly parametric. 
%While these parametric techniques do not allow the data to ``speak for itself", the ensuing data analysis results typically are readily interpretable, providing important subject area insights.
%At the other extreme are black box type methods of machine learning.  
%While these methods are incredibly successful at predicting how processes evolve, they provide little structured insight into the driving mechanisms of the process.
%Functional Data Analysis (FDA) techniques lie somewhere in between these extremes: the output of an analysis provides visual tools for understanding processes and, often, some dimension reduction to allow for easier exploration.  
%The development of structured FDA techniques which incorporate more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).

Biologging data are sometimes modelled as a set of curves that can be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}). In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve where metabolic activity is a function of time. The dive profile of a marine mammal can also be viewed as a curve. \citet{Fu:2017} show that estimating the amplitude and phase variation of elephant seal (\textit{Mirounga leonina}) dive curves is useful to classify dive types, although they ignore sequential dependence between dives. 
%However, now many such animals are tracked with high-frequency tags over a period of several dives, providing information at different time scales. 
Similarly, our case study views the dive profile of a killer whale (\textit{Orcinus orca}) as a sequence of curves where three orthogonal axes of acceleration are each functions of time.

FDA was originally developed for curves that can be considered independent replicates, but the structure of time-series data is often complicated by additional dependencies. In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017,Adam:2019}, as is the case for our case study data. Namely, the dive profile (Figure \ref{fig:data}a) of a killer whale from the threatened Northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence \textit{between} dives, as dives of various time durations and depths appear to be clustered together in time. Previous studies have categorized these dives into different types where each dive type is associated with a different behaviour, including resting, travelling, and foraging \citep{Tennessen:2019b}.
Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence \textit{within} dives, as the accelerometer data within each dive indicate bouts of short-term periodicity. The combination of apparent behavioural changes combined with periodic behaviour makes this accelerometer data difficult to model using simple stochastic processes. Another example of behaviourally-dependent periodicity is in the field of machine health, where \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks. The raw acceleration data exhibit extreme periodicity (in the form of vibration) that depends upon the behaviour of the train as the load and speed of the train change.

FDA literature provides models that account for some between-curve dependence, but these methods are not adequate when this dependence is in time. For example, multilevel models incorporate dependence via random effects. In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors use a multilevel approach since each child yields two daily activity curves; the model includes a random effect for ``child''. In a classic paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles. More recent work on multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}. However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types. 
In addition to multilevel models, FDA researchers have used functional time series to model dependence in sequences of curves. Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}. While functional time series methods can in theory be applied to biologging data such as ours, they do not account for sequences of curves that switch between a discrete number of types over time.

Traditional FDA techniques similarly fall short when modelling complicated within-curve acceleration data. In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function. \citet{Yao:2005} use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis. 
%In terms of applications, \citet{Gruen:2017} and \citet{Xiao:2014} investigate degenerative joint disease in cats and circadian rhythms in aging patients, respectively. 
Alternatively, within-curve structure is sometimes generated via random regression. For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves. However, radically changing behaviour on short time-scales similar to our dive sequence is challenging to model with these classic FDA techniques.

In light of these challenges, we turn to vast field of methodology for the analysis of animal movement \citep{Hooten:2017}, where one of the most prevalent techniques in the literature of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}. HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of the animal from movement observations. While ubiquitous in the ecology literature, HMMs have seen little use in non-parametric functional modelling, with a few notable exceptions. In particular, \citet{Langrock:2018} take a non-parametric approach to model the distribution of HMM observations in terms of B-splines. This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments. However, this approach does not account for certain types of time correlation within or between the time series themselves. We focus on the time series themselves as functions here and show how HHMs can provide a new tool to the FDA toolbox.

%
%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another when conditioned on the underlying behavioural process.
%However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
%Several publications have dealt with this issue in the past, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
%The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
%while maintaining the structure needed to run the forward algorithm. 
%It also only adds one additional parameter per possible hidden state.

Under a traditional HMM, observations are independent given the underlying hidden state process. However, this assumption is often violated in real-world processes, especially when observations are taken at high frequencies. Several solutions have been proposed in the ecology literature, such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. The CarHMM in particular explicitly models auto-correlation into an HMM while adding minimal complexity to the base structure of an HMM. Animal movement modellers have also used hierarchical hidden Markov models (HHMMs) to model simultaneous behavioural processes that occur at different time scales \citep{Barajas:2017,Adam:2019}. We borrow from ideas from both the CarHMM and the HHMM here.

Although they are useful in many situations, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated structures at short time scales. The HHMM of \citet{Adam:2019} fails to capture the fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}). \citet{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM to account for periodic behaviour in the context of describing daily behavioural cycles of marine mammals. In addition, Fourier analysis has previously been used on accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}. Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for periodic structures in fine-scale accelerometer data.

In this paper, we combine several existing methods from the ecology literature in novel ways to account for complex temporally dependent functional data. The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models and thus extend the methods available for the analysis of functional data.
%This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. 
%We describe a general procedure that can be used to extract features from highly structured fine-scale behaviours that otherwise could not be modelled with existing HMM models. 
%In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. 
%We then perform a simulation study to compare the performance of each existing model with ours in a controlled setting.
We first describe HMMs as well as to two variants (CarHMMs and HHMMs) and discuss how Fourier analysis can handle fine-scale dependence structures. We also show how these methods can be combined to analyze increasingly complex data. In section 3, we construct several candidate models and fit them to the data from Figure \ref{fig:data}. Section 4 details a simulation study based on these candidate models, and in section 5 we discuss our results.

>>>>

Biologging technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}. In particular, the study of animal movement has been transformed by tracking devices that collect biologging data \citep{Borger:2020}. Tags can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. These data contain a wealth of knowledge about human and animal behaviour, but parsing these large data sets is a challenge for statisticians.

Biologging data are sometimes modelled as a set of curves that can be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}). For example, \cite{Morris:2007} view a child's activity level throughout the day as a curve where metabolic activity is a function of time. \citet{Fu:2017} also show that estimating the amplitude and phase variation of elephant seal (\textit{Mirounga leonina}) dive curves is useful to classify dive types, although they ignore sequential dependence between dives. Similarly, our case study views the dive profile of a killer whale (\textit{Orcinus orca}) as a sequence of curves where three orthogonal axes of acceleration are each functions of time.

FDA was originally developed for curves that can be considered independent replicates, but the structure of time-series data is often complicated by additional dependencies. In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017,Adam:2019}, as is the case for our case study data. Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time. Previous studies have categorized these dives into different types where each dive type is associated with a different behaviour, including resting, travelling, and foraging \citep{Tennessen:2019b}.
Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence within dives, as the accelerometer data within each dive indicate bouts of short-term periodicity. Behaviourally dependent periodicity is not unique to the biologging data- in the field of machine health, \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks. The raw acceleration data exhibit extreme periodicity (in the form of vibration) that depends upon the behaviour of the train as the load and speed of the train change.

FDA literature provides models that account for some between-curve dependence, but these methods are not adequate when modelling certain types of time dependencies, including that of Figure \ref{fig:data}. For example, multilevel models incorporate dependence via random effects. In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors use a multilevel approach with a random effects to model variation between children. In a classic paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles. More recent work on multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}. However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types. 
In addition to multilevel models, FDA researchers have used functional time series to model dependence in sequences of curves. Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}. While functional time series methods can in theory be applied to biologging data such as ours, they do not account for sequences of curves that switch between a discrete number of types over time.

Traditional FDA techniques similarly fall short when modelling complicated within-curve acceleration data. In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function. \citet{Yao:2005} use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis. Alternatively, within-curve structure is sometimes generated via random regression. For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves. However, The combination of apparent behavioural changes combined with periodic behaviour makes accelerometer data such as ours difficult to model with these classic FDA techniques.

In light of these challenges, we turn to vast field of methodology for the analysis of animal movement \citep{Hooten:2017}, where one of the most prevalent techniques in the literature of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}. HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from movement observations. While ubiquitous in the ecology literature, HMMs have seen little use in non-parametric functional modelling, with a few notable exceptions. In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations in terms of B-splines. 
%This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments. 
However, this approach does not account for certain types of time correlation within or between the time series themselves, which is the focus of our efforts here.

While useful, traditional HMMs are also not sufficient to model accelerometer data such as ours for three primary reasons.
Firstly, subsequent observations are assumed to be independent given the underlying hidden state process, but this is often not the case when observations are taken at high frequencies. Several solutions have been proposed in the ecology literature, such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. The CarHMM in particular explicitly models auto-correlation into an HMM while adding minimal complexity to the base structure of an HMM.

In addition, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales, including our example of both between- and within-curve temporal dependencies. To deal with this issue, Animal movement modellers have used hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}. In particular, HHMMs treat both scales as HMMs where the behaviour of the fine-scale process depends on the state of the coarse-scale process.

Finally, although they are useful in many situations, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated structures at short time scales. The HHMM of \citet{Adam:2019} fails to capture the fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}). \citet{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM to account for periodic behaviour in the context of describing daily behavioural cycles of marine mammals. In addition, Fourier analysis has previously been used on accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}. Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for periodic structures in fine-scale accelerometer data.

While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together. In this paper, we combine several existing methods from the ecology literature in novel ways to account for complex temporally dependent functional data. The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models and thus extend the methods available for the analysis of functional data. We first describe HMMs as well as to two variants (CarHMMs and HHMMs) and discuss how Fourier analysis can handle fine-scale dependence structures. We also show how these methods can be combined to analyze increasingly complex data. In Section 3, we construct several candidate models and fit them to the data from Figure \ref{fig:data}. Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.
